\(Q\)-Boson model and relations with integrable hierarchies

### 2024-04-23 draft.tex
improve tau function section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

With these expressions, one can finally build the Bethe states
\begin{equation}
  \ket{\Psi(y_1, \dots, y_N)} = \prod_{j=1}^N \mathbb{B}(y_j) \ket{\bm{0}} \qquad 
  \bra{\Psi(y_1, \dots, y_N)} = \bra{\bm{0}}\prod_{j=1}^N \mathbb{C}(y_j) 
\end{equation}
where \(\mathbb{B}(x) = x^{M/2} B(x)\) and \(\mathbb{C}(x) = x^{M/2} C(1/x)\).

When the set \(\{ y_j \ | \ j =1, \dots , N\}\) satisfies the Bethe equations
\begin{equation}
\label{eq:bethe_eq}
  y^{N + M}_i = (-1)^{N-1} \prod_{\substack{j = 1 \\ j \neq i}}^N y_j\; , \qquad i = 1, \dots, N\; , 
\end{equation}
we say that the Bethe states are \emph{on-shell}; otherwise, we have
\emph{off-shell} states. In what follows, we will only consider
\emph{off-shell} states.

@@

\subsection{q-Bosons}
The Hamiltonian for the q-boson model is
given by
\begin{equation}
  \mathcal{H} = -\frac{1}{2} \sum_{n=0}^M
  \left(b_n^\dagger b_{n+1} + b_n b_{n+1}^\dagger \right) + \bar{\mathcal{N}}\; ,
\end{equation}
where \(\bar{\mathcal{N}} = \sum_{i=0}^M \mathcal{N}_i\), and as in
the case of the phase mode, we impose periodic boundary conditions
\(b_{M+1} = b_0\) and \(b_{M+1}^\dagger = b_0^\dagger\). The set of
operators \(\{b_i, b^\dagger,\mathcal{N}_i\}_{i=0}^M\) form the
\(M+1\) copies of independent q-boson algebras:
\begin{equation}
[\mathcal{N}_i, b_j^\dagger]=\delta_{i,j} b_i^\dagger\; , \quad 
[\mathcal{N}_i, b_j]=-\delta_{i,j}b_i\; , \quad
[b_i, b_j^\dagger]= \delta_{i,j} q^{-2\mathcal{N}_i}  \equiv \delta_{i,j} Q^{\mathcal{N}_i}\; , 
\end{equation}
where we write the deformation parameter as \(Q = q^{-2}\). 
In the limit \(Q\to 1\), the q-bosons are ordinary bosons, and 
the limit \(Q\to 0\) (\(q\to \infty\)) gives the phase model we have discussed above.

\subsubsection{Representation}
We build the representation of the q-boson algebra with some modifications of the phase 
model we have just discussed. Let us define the i\(^{th}\)-vacuum \(\ket{0}_i\) by
\(b_i\ket{0}_i = 0\). The representation of this Hilbert space, denoted by
\(\mathcal{F}_i^Q\), is given by \(\ket{n_i}_i \propto b_i^\dagger \ket{0}_i\).

Given the vacuum \(\ket{\bm{0}} = \ket{0}_0\otimes \ket{0}_1
\otimes \cdots \otimes  \ket{0}_M\),
the Fock space is defined as 
\begin{equation}
  \mathcal{F}_{Q} = \bigotimes_{i=0}^M \mathcal{F}_i^Q = 
  \left\{\ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots
  \otimes \ket{n_M} \ | \ n_i \in \mathbb{N} \right\}\; .
\end{equation}
where the actions of the operators \(\{b_i, b_i^\dagger\}\) are given by 
the following relations
\begin{subequations}
\begin{alignat}{3}
    & b_0\ket{n_0} =(1 - \delta_{0, n_0}) \frac{(1 - Q^{n_0})}{(1 - Q)^{1/2}}\ket{n_0 - 1}
    &\hspace{0.5cm}& b_0^\dagger \ket{n_0} =  \frac{1}{(1 - Q)^{1/2}}\ket{n_0 + 1}  \\
    & b_i\ket{n_i} = \frac{(1 - \delta_{0, n_i})}{(1 - Q)^{1/2}}\ket{n_i - 1}
    && b_i^\dagger \ket{n_0} =  \frac{(1 - Q^{n_i+1})}{(1 - Q)^{1/2}}\ket{n_0 + 1}\quad i\neq 0\; .
\end{alignat}
\end{subequations}
Observe that the modes \(\{b_0, b_0^\dagger \}\) satisfies a slightly
different relation.

Therefore, the states \(\ket{\vec{n}} \in \mathcal{F}_Q\) are defined as 
\begin{equation}
  \ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots \otimes \ket{n_M} 
 : =  (b_0^\dagger)^{n_0} (b_1^\dagger)^{n_1} \cdots  (b_M^\dagger)^{n_M} \ket{\bm{0}} \; ,
\end{equation}
where we write \(\phi_i^\dagger \equiv \bm{1} \otimes  \cdots \otimes
\phi_i^\dagger \otimes \cdots \otimes \bm{1}\).
Moreover, the norm of these states satisfy 
\begin{equation}
  \bracket{\vec{n}}{\vec{m}} = \frac{[m_0]!}{\prod_{i=1}^M[m_i]!} \delta_{\vec{n}\vec{m}}\qquad 
    [n]! =
    \left\{
    \begin{array}{ll}
    \prod_{i=1}^n(1 - Q^i) & \textrm{if}\ \ n\neq 0 \\
    1 & \textrm{otherwise}
  \end{array}\right.\; .
\end{equation}
As before, we can associate to a given states \(\ket{\vec{n}} =
\ket{n_0, n_1, \dots, n_M}\), a partition -- or Young diagram --
\(\lambda = (1^{n_1} 2^{n_2} \cdots M^{n_M})\), but now we assume that 
there is a proportionaly  factor relating these two objects, 
\begin{equation}
  \ket{\lambda}_Q  = b_\lambda(Q) \ket{\vec{n}}  \ \; , \qquad b_\lambda(Q) = \prod_i [p_i(\lambda)]! 
\end{equation}
where \(p_i(\lambda)\) denotes the number of parts of size \(i\) in
the partition. These partition states satisfy \(
\bracket{\lambda}{\mu}_Q = b_\lambda(Q) \delta_{\lambda,\mu}\). Once
again, we see that the correspondence is not unique since the number
of oscilators at the site \(i=0\) are completely ignored in the
partition notation. And finally, if the number of particles is fixed,
then \(n_0 = N - \ell(\lambda)\).

\subsubsection{Bethe Ansatz}
The \(L\)-operator for the q-boson is given by
\begin{equation}
  L_{an}(x, Q) =
  \begin{pmatrix}
    x^{-1/2} & (1 - Q)^{\frac{1}{2}} b_n^\dagger \\ (1 - Q)^{\frac{1}{2}} b_n & x^{1/2}
  \end{pmatrix}_a\; .
\end{equation}
The monodromy matrix is 
\begin{equation}
  T_a(x,Q) = L_{am}(x, Q)  \dots  L_{a0}(x, Q) = 
  \begin{pmatrix}
    A(x, Q) & B(x, Q) \\ C(x, Q) & D(x, Q)
  \end{pmatrix}_a\; .
\end{equation}

The eigenstates of the Hamiltonian are of the form
\begin{equation}
  \ket{\Psi(y_1, \dots, y_N; Q)} = \prod_{j=1}^N \mathbb{B}(y_j, Q) \ket{\bm{0}}\qquad 
  \bra{\Psi(y_1, \dots, y_N; Q)} = \bra{\bm{0}}\prod_{j=1}^N \mathbb{C}(y_j, Q) \; .
\end{equation}
where \(\mathbb{B}(x, Q) = x^{M/2} B(x, Q)\) and \(\mathbb{C}(x, Q) =
x^{M/2} C(1/x, Q)\).  When the set of parameters \(\{ y_j \ | \ j =1, \dots , N\}\)
satisfies the Bethe equations
\begin{equation}
  y^{N + M}_i =\prod_{\substack{j = 1 \\ j \neq i}}^N\frac{Q y_i - y_j}{y_i - Q y_j}\; , \qquad i = 1, \dots, N\; , 
\end{equation}
we say that the states \(\ket{\Psi(y_1, \dots, y_N; Q)}\) are \emph{on-shell} Bethe states, otherwise, 
we have \emph{off-shell} Bethe states.

@@

This section discusses the presence of integrable hierarchies in the phase
model.  We argue that it is also a Toda hierachy tau function, and we
discuss some consequences of this fact, in particular, its connection
with a matrix model. Moreover, we also point the presence of
correlation functions in this model that also satisfy the KP hierarchy
equations. We finally argue how these results also imply that these
scalar products are also KP hierarchy tau functions, what is expected from 
the results of Wheeler~\cite{Wheeler:2010vmq}.

Bogoliubov~\cite{Bogoliubov2005}, has shown that the scalar product of
two vectors in the \(N\)-particles sector in a chain of length \(M+1\)
is
\begin{equation}
\begin{split}
\label{eq:scalar}
  \mathcal{I}(N, M|\bm{x}, \bm{y}) & =
  \bra{\bm{0}} \prod_{j=1}^N \mathbb{C}(x_j) \prod_{j=1}^N \mathbb{B}(y_j) \ket{\bm{0}} \\ 
 & = \frac{ \det H(\bm{x},\bm{y})}{ \prod_{i<j}(x_i - x_j)(y_i - y_j)} \; ,
\end{split}
\end{equation}
where \(H\) is an \(N\times N\) matrix with components
\begin{equation}
\label{eq:h-matrix}
  H_{ij} = H(x_i, y_j) 
  =\frac{1 - (x_i y_j)^{ M + N}}{1 - x_i y_j }\; .
\end{equation}

@@

\subsection{Toda tau functions}
In this section, we argue that the scalar product defined above is a tau function 
of the Toda hierarchy. Let us first write the function \(H(z,w)\) as the geometric sum 
\begin{equation}
\label{eq:h-exp}
  H(z,w) = \frac{1 - (zw)^{M+N}}{1 - zw} = \sum_{k=1}^{M+N} (zw)^{k-1} \; , 
\end{equation}
in such a way that 
we can write
\begin{equation}
  \det_{i,j} \left(H(x_i, y_j)\right) = \det_{i,j} \left( \sum_{k=1}^{M+N} x_i^{k-1} y_j^{k-1}\right)
\end{equation}
and we can thing of this expression as the product of a \(N\times (N+M)\) matrix \(\mathcal{X}\)
and another \((M + N)\times N\) matrix \(\mathcal{Y}\), given by
\begin{equation}
  \mathcal{X} = 
  \begin{pmatrix}
  x_1^0 & x_1^1 & \dots & x_1^{M+N-1} \\  
  x_2^0 & x_2^1 & \dots & x_2^{M+N-1} \\  
  \vdots \\
  x_N^0 & x_N^1 & \dots & x_N^{M+N-1} 
  \end{pmatrix}\quad \textrm{and} \quad 
  \mathcal{Y} = 
  \begin{pmatrix}
  y_1^0 & y_2^0 & \dots & y_N^0 \\  
  y_1^1 & y_2^1 & \dots & y_N^1 \\  
  \vdots & \vdots & & \vdots \\
  y_1^{N+M-1} & y_2^{M+N-1} & \dots & y_N^{M+N-1} 
  \end{pmatrix}
\end{equation}
therefore 
\begin{equation}
  \det_{i,j} \left(H(x_i, y_j)\right) =  \det_{i,j} \left( \mathcal{X}\mathcal{Y}\right)
= \sum_{0 \leq \ell_{N} \leq \dots \leq \ell_1\leq N+M }  \det_{ik}(x_i^{\ell_k})  \det_{ik}(y_j^{\ell_k})\; .
\end{equation}
If we now define \(\ell_j = \lambda_k - k + N\), and
using~(\ref{eq:scalar}), we have
\begin{equation}
\label{eq:scalar_exp}
\mathcal{I}(N, M|\bm{x}, \bm{y}) = \sum_{\lambda\subseteq [N,M]}
\frac{\det_{ik}(x_i^{\lambda_k - k + N})}{\Delta(\bm{x})} \frac{ \det_{ik}(y_j^{\lambda_k - k + N})}{\Delta(\bm{y})}
= \sum_{\lambda\subseteq [N,M]} s_\lambda(\bm{x}) s_\lambda(\bm{y}) \; .
\end{equation}
where \(\Delta(\bm{x})\) and \(\Delta(\bm{x})\) are Vandermonde
determinants. This formula agrees with the Schur expansion defined
in~\cite{Bogoliubov2005}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

With these expressions, one can finally build the Bethe states
\begin{equation}
  \ket{\Psi(y_1, \dots, y_N)} = \prod_{j=1}^N \mathbb{B}(y_j) \ket{\bm{0}} \qquad 
  \bra{\Psi(y_1, \dots, y_N)} = \bra{\bm{0}}\prod_{j=1}^N \mathbb{C}(y_j) 
\end{equation}
where \(\mathbb{B}(y) = y^{M/2} B(y)\) and \(\mathbb{C}(y) = y^{M/2} C(1/y)\).

When the coordinates \(\{ y_j \ | \ j =1, \dots , N\}\) satisfy the Bethe equations
\begin{equation}
\label{eq:bethe_eq}
  y^{N + M}_i = (-1)^{N-1} \prod_{\substack{j = 1 \\ j \neq i}}^N y_j\; , \qquad i = 1, \dots, N\; , 
\end{equation}
we say that the Bethe states are \emph{on-shell}; otherwise, we have
\emph{off-shell} states. In what follows, we will only consider
\emph{off-shell} states.

@@

\subsection{q-Bosons}
The set of operators \(\{b_i, b^\dagger,\mathcal{N}_i\}_{i=0}^M\) constitute
\(M+1\) independent q-boson algebras defined by
\begin{equation}
[\mathcal{N}_i, b_j^\dagger]=\delta_{i,j} b_i^\dagger\; , \quad 
[\mathcal{N}_i, b_j]=-\delta_{i,j}b_i\; , \quad
[b_i, b_j^\dagger]= \delta_{i,j} q^{-2\mathcal{N}_i}  \equiv \delta_{i,j} Q^{\mathcal{N}_i}\; , 
\end{equation}
where we denote the deformation parameter as \(Q = q^{-2}\).

The q-boson model is characterized by its Hamiltonian
\begin{equation}
  \mathcal{H} = -\frac{1}{2} \sum_{n=0}^M
  \left(b_n^\dagger b_{n+1} + b_n b_{n+1}^\dagger \right) + \bar{\mathcal{N}}\; ,
\end{equation}
where \(\bar{\mathcal{N}} = \sum_{i=0}^M \mathcal{N}_i\), and we also
impose periodic boundary conditions \(b_{M+1} = b_0\) and
\(b_{M+1}^\dagger = b_0^\dagger\).

In the limit \(Q \to 1\), the q-bosons behave as ordinary bosons,
while the limit \(Q \to 0\) (\(q \to \infty\)) corresponds to the
phase model discussed earlier.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Representation}
We construct the representation of the q-boson algebra with
modifications of the phase model discussed earlier. Let us define the
\(i\)-th vacuum \(\ket{0}_i\) such that \(b_i\ket{0}_i = 0\). The
representation of this Hilbert space, denoted by \(\mathcal{F}_i^Q\),
is given by \(\ket{n_i}_i \propto b_i^\dagger \ket{0}_i\).

Given the vacuum \(\ket{\bm{0}} = \ket{0}_0\otimes \ket{0}_1
\otimes \cdots \otimes  \ket{0}_M\),
the Fock space is defined as 
\begin{equation}
  \mathcal{F}_{Q} = \bigotimes_{i=0}^M \mathcal{F}_i^Q = 
  \left\{\ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots
  \otimes \ket{n_M} \ | \ n_i \in \mathbb{N} \right\}\; ,
\end{equation}
where the actions of the operators \(\{b_i, b_i^\dagger\}\) are given
by the following relations
\begin{subequations}
\begin{alignat}{3}
    & b_0\ket{n_0} =(1 - \delta_{0, n_0}) \frac{(1 - Q^{n_0})}{(1 - Q)^{1/2}}\ket{n_0 - 1}
    &\hspace{0.5cm}& b_0^\dagger \ket{n_0} =  \frac{1}{(1 - Q)^{1/2}}\ket{n_0 + 1}  \\
    & b_i\ket{n_i} = \frac{(1 - \delta_{0, n_i})}{(1 - Q)^{1/2}}\ket{n_i - 1}
    && b_i^\dagger \ket{n_0} =  \frac{(1 - Q^{n_i+1})}{(1 - Q)^{1/2}}\ket{n_0 + 1}\quad i\neq 0\; .
\end{alignat}
\end{subequations}

Consequently, the states \(\ket{\vec{n}} \in \mathcal{F}_Q\) are
\begin{equation}
  \ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots \otimes \ket{n_M} 
 : =  (b_0^\dagger)^{n_0} (b_1^\dagger)^{n_1} \cdots  (b_M^\dagger)^{n_M} \ket{\bm{0}} \; ,
\end{equation}
where we write \(b_i^\dagger \equiv \bm{1} \otimes  \cdots \otimes
b_i^\dagger \otimes \cdots \otimes \bm{1}\).
Moreover, this space has an inner product that satisfies
\begin{equation}
  \bracket{\vec{n}}{\vec{m}} = \frac{[m_0]!}{\prod_{i=1}^M[m_i]!} \delta_{\vec{n}\vec{m}}\qquad 
    [n]! =
    \left\{
    \begin{array}{ll}
    \prod_{i=1}^n(1 - Q^i) & \textrm{if}\ \ n\neq 0 \\
    1 & \textrm{otherwise}
  \end{array}\right.\; .
\end{equation}

Similar to the phase model, we can associate to a given states
\(\ket{\vec{n}} = \ket{n_0, n_1, \dots, n_M}\), a Young diagram
\(\lambda = (1^{n_1} 2^{n_2} \cdots M^{n_M})\). It is useful to define
a proportionaly factor relating these two objects
\begin{equation}
  \ket{\lambda}_Q  = b_\lambda(Q) \ket{\vec{n}}  \ \; , \qquad b_\lambda(Q) = \prod_i [p_i(\lambda)]! 
\end{equation}
where \(p_i(\lambda)\) denotes the number of parts of size \(i\) in
the partition. These partition states satisfy
\(\bracket{\lambda}{\mu}_Q = b_\lambda(Q) \delta_{\lambda,\mu}\). Once
again, we see that the correspondence is not unique, as the number of
oscillators at site \(i=0\) is completely ignored in the partition
notation. Finally, if the number of particles is fixed, then \(n_0 = N
- \ell(\lambda)\).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Bethe Ansatz}
The \(L\)-operator for the q-boson is given by
\begin{equation}
  L_{an}(x, Q) =
  \begin{pmatrix}
    x^{-1/2} & (1 - Q)^{\frac{1}{2}} b_n^\dagger \\ (1 - Q)^{\frac{1}{2}} b_n & x^{1/2}
  \end{pmatrix}_a\; ,
\end{equation}
and the monodromy matrix is 
\begin{equation}
  T_a(x,Q) = L_{am}(x, Q)  \dots  L_{a0}(x, Q) = 
  \begin{pmatrix}
    A(x, Q) & B(x, Q) \\ C(x, Q) & D(x, Q)
  \end{pmatrix}_a\; .
\end{equation}

As before, the eigenstates of the Hamiltonian have the form
\begin{equation}
  \ket{\Psi(y_1, \dots, y_N; Q)} = \prod_{j=1}^N \mathbb{B}(y_j, Q) \ket{\bm{0}}\qquad 
  \bra{\Psi(y_1, \dots, y_N; Q)} = \bra{\bm{0}}\prod_{j=1}^N \mathbb{C}(y_j, Q) \; .
\end{equation}
where \(\mathbb{B}(y, Q) = y^{M/2} B(y, Q)\) and \(\mathbb{C}(y, Q) =
y^{M/2} C(1/y, Q)\). When the parameters \(\{ y_j \ | \ j =1, \dots , N\}\)
satisfy the Bethe equations given by
\begin{equation}
  y^{N + M}_i =\prod_{\substack{j = 1 \\ j \neq i}}^N\frac{Q y_i - y_j}{y_i - Q y_j}\; , \qquad i = 1, \dots, N\; , 
\end{equation}
we have \emph{on-shell} Bethe states; otherwise, they are \emph{off-shell} states.

This section explores the presence of integrable hierarchies in the
phase model. We demonstrate its correspondence to a Toda hierarchy tau
function and discuss some implications, particularly its connection to
a matrix model. Additionally, we highlight the existence of
correlation functions in this model that satisfy the KP hierarchy
equations. Finally, we argue that these results imply that these
scalar products are also KP hierarchy tau functions, aligning with the
findings of Wheeler~\cite{Wheeler:2010vmq}.

Bogoliubov~\cite{Bogoliubov2005} has demonstrated that the scalar
product of two vectors in the \(N\)-particle sector of a chain with
length \(M+1\) is
\begin{equation}
\begin{split}
\label{eq:scalar}
  \mathcal{I}(N, M|\bm{x}, \bm{y}) & =
  \bra{\bm{0}} \prod_{i=1}^N \mathbb{C}(x_i) \prod_{j=1}^N \mathbb{B}(y_j) \ket{\bm{0}} \\ 
 & = \frac{ \det H(\bm{x},\bm{y})}{ \prod_{i<j}(x_i - x_j)(y_i - y_j)} \; ,
\end{split}
\end{equation}
where \(H\equiv [H_{ij}]_{i,j=1}^N\) is an \(N\times N\) matrix with components
\begin{equation}
\label{eq:h-matrix}
  H_{ij} = H(x_i, y_j) 
  =\frac{1 - (x_i y_j)^{ M + N}}{1 - x_i y_j }\; .
\end{equation}

@@

\subsection{Toda tau functions}
We now argue that the scalar product defined above is a tau function
of the Toda hierarchy. Let us first write the function \(H(z,w)\) as
the geometric sum
\begin{equation}
\label{eq:h-exp}
  H(z,w) = \frac{1 - (zw)^{M+N}}{1 - zw} = \sum_{k=1}^{M+N} (zw)^{k-1} \; .
\end{equation}
Hence, we express the determinant of the \(H\) matrix as
\begin{equation}
  \det_{i,j} \left(H(x_i, y_j)\right) = \det_{i,j} \left( \sum_{k=1}^{M+N} x_i^{k-1} y_j^{k-1}\right)
\end{equation}
Furthermore, we can interpret this expression as the result of
multiplying an \(N\times (N+M)\) matrix \(\mathcal{X}\) by another
\((M + N)\times N\) matrix \(\mathcal{Y}\), which are given by
\begin{equation}
  \mathcal{X} = 
  \begin{pmatrix}
  x_1^0 & x_1^1 & \dots & x_1^{M+N-1} \\  
  x_2^0 & x_2^1 & \dots & x_2^{M+N-1} \\  
  \vdots \\
  x_N^0 & x_N^1 & \dots & x_N^{M+N-1} 
  \end{pmatrix}\quad \textrm{and} \quad 
  \mathcal{Y} = 
  \begin{pmatrix}
  y_1^0 & y_2^0 & \dots & y_N^0 \\  
  y_1^1 & y_2^1 & \dots & y_N^1 \\  
  \vdots & \vdots & & \vdots \\
  y_1^{N+M-1} & y_2^{M+N-1} & \dots & y_N^{M+N-1}
  \end{pmatrix}\; ,
\end{equation}
therefore 
\begin{equation}
  \det_{i,j} \left(H(x_i, y_j)\right)
  \equiv  \det_{i,j} \left( \mathcal{X}\mathcal{Y}\right)
  = \sum_{0 \leq \ell_{N} \leq \dots \leq \ell_1\leq N+M } \det_{ik}(x_i^{\ell_k}) \det_{ik}(y_j^{\ell_k})\; .
\end{equation}
If we now define \(\ell_j = \lambda_k - k + N\), and
using~(\ref{eq:scalar}), we have
\begin{equation}
\label{eq:scalar_exp}
\mathcal{I}(N, M|\bm{x}, \bm{y}) = \sum_{\lambda\subseteq [N,M]}
\frac{\det_{ik}(x_i^{\lambda_k - k + N})}{\Delta(\bm{x})} \frac{ \det_{ik}(y_j^{\lambda_k - k + N})}{\Delta(\bm{y})}
= \sum_{\lambda\subseteq [N,M]} s_\lambda(\bm{x}) s_\lambda(\bm{y}) \; .
\end{equation}
where \(\Delta(\bm{x})\) and \(\Delta(\bm{x})\) are Vandermonde
determinants. This formula agrees with the Schur expansion defined
in~\cite{Bogoliubov2005}.

Define two sets of Miwa coordinates \(\bm{t} = (t_1, t_2, \dots)\) and
\(\bm{t}' = (t'_{-1}, t'_{-2}, \dots)\) by
\begin{equation}
  t_p = \frac{1}{p}\sum_{j=1}^N x_j^p\qquad 
  t'_{-p} = \frac{1}{p}\sum_{j=1}^N y_j^p\; .
\end{equation}
One can write the inner product in terms of these coordinates, that is
\begin{equation}
  \mathcal{I}(N, M|\bm{t}, \bm{t}') = \sum_{\lambda\subseteq [N,M]} s_\lambda(\bm{t}) s_\lambda(\bm{t}') \; .
\end{equation}
This expression is known to be a tau function for \(M, N\to \infty\).
As such, utilizing the free fermions representation~\cite{Alexandrov:2012tr}, it can be written as
\begin{equation}
  \lim_{M,N\to \infty}\mathcal{I}(N, M|\bm{t}, \bm{t}')
 = \bra{\bm{0}} e^{\bm{J}_+(\bm{t})} e^{-\bm{J}_-(\bm{t}')} \ket{\bm{0}}\; .
\end{equation}
It is a tau function of the Toda hierarchy with trivial element \(\mathbb{1} \in GL(\infty)\).
It is nothing but the Cauchy's identity
\begin{equation}
  \sum_{\lambda } s_{\lambda}(\bm{t}) s_{\lambda}(\bm{t}')
    = \exp \left( \sum_{m\geq1} m t_m t'_{-m} \right) \; .
\end{equation}

Bringing all these facts together, the truncation for finite \(M\) and
\(N\) also yields tau functions of the Toda hierarchy. More
specifically, according to~\cite{Alexandrov:2012tr, Kharchev:1991gd,
  Zabrodin:2010ii}, the truncation of the tau function corresponds to
the inclusion of a projection operator \(\mathrm{P}^+\) in the
expectation value of the tau function written in the fermionic
representation.

As a final remark, we can also write
\begin{equation}
  H(z,w) = \sum_{k=0}^{M+N} h_k(zw)^{k-1} \; , 
\end{equation}
with \(h_k=1\) if \(k\in [0, M+N-1]\) and \(0\) otherwise. In this
case, one can define a diagonal \((N+M)\times (N+M)\) matrix
\(\mathcal{H} = \textrm{diag}(h_0, \dots, h_{M+N-1})\). Consequently,
if we repeat the arguments above, we find
\begin{equation}
\mathcal{I}(N, M|\bm{t}, \bm{t}') = \sum_{\lambda} h_\lambda s_\lambda(\bm{t}) s_\lambda(\bm{t}') \; .
\end{equation}
Here, \(h_\lambda\) is equal to \(1\) if \(\lambda \in [N,M]\), and it
is zero otherwise.

In this case, we find that this tau function is a trivial example of
the tau functions considered in~\cite{orlov:2001}. We anticipate that
in the analysis of more general correlation functions, the diagonal
terms \(h_\lambda\) will be more interesting. We will revisit this
discussion soon.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Matrix Models}
It is also interesting to note the particular case when \(M\to
\infty\), but with a finite number of particles \(N\). In this case,
the partitions \(\lambda \in [N, \infty]\) satisfy the condition
\(\ell(\lambda) \leq N\). Therefore, the scalar
product~(\ref{eq:scalar_exp}) becomes
\begin{equation}
  \mathcal{I}_N(\bm{t}, \bm{t}')\equiv 
  \lim_{M\to \infty}\mathcal{I}(N, M|\bm{t}, \bm{t}')
  = \sum_{\substack{\lambda \\ \ell(\lambda) \leq N}} s_\lambda(\bm{t})  s_\lambda(\bm{t}') \; .
\end{equation}
From~\cite{Zabrodin:2010ii}, we know that this expression can be
written as the following integral
\begin{equation}
  \mathcal{I}_N(\bm{t}, \bm{t}') =
  \frac{1}{N!} \prod_{\ell=1}^N \oint_{\Gamma_\ell} \frac{dz_\ell}{2 \pi i z_\ell}
  e^{\xi(\bm{t}, z_\ell) - \xi(\bm{t}', z_{\ell}^{-1})} \Delta(z)\Delta(z^{-1})\; ,
\end{equation}
where
\begin{equation}
\xi(\bm{t}, z) = \sum_{k\geq 0} t_k z^k \qquad
\xi(\bm{t}', 1/z) = \sum_{k\geq 0} t'_{-k} z^{-k} \; . 
\end{equation}
See also~\cite{Kharchev:1991gd} for a detailed proof of this relation,
and~\cite{Orlov:2005} for other details.

From the results of~\cite{Zabrodin:2010ii}, see citations therein, we
have an interesting consequence of this representation. Impose the
Bethe equations~(\ref{eq:bethe_eq}) to one set of variables, say \(x_j
= e^{-ip_j} \in \mathbb{S}^1\). Additionally, let us set \(\bm{t}' = -
\bm{t}^\star\). In this particular case, we have
\begin{equation}
\xi(\bm{t}, z) - \xi(\bm{t}', 1/z)  = 2 \ \textrm{Re}\left(\sum_k t_k z^k\right)\; .
\end{equation}
Then, the phase model is equivalent to an ensemble of \(N\) 2D Coulomb
particles on a circle. In this case, we find that the quantities
\(\lambda_{\ell}\) are eigenvalues of a matrix \(U\).

Furthermore, according to Zabrodin~\cite{Zabrodin:2010ii}, see
citations therein, we also know that under the rescaling \(t_k \to
T_k/ \hbar\), \(t'_k \to T_{-k}/ \hbar\) and \(N = T_0/ \hbar\), we
obtain the dispersion limit tau function
\begin{equation}
  F_0(\bm{T}) = \log S_{T_0}(\bm{T}, \bm{T}') + \mathcal{O}(\hbar)\; ,
\end{equation}
that is a free energy, from the viewpoit of the matrix integral
partition function.

It remains unclear how one can use this fact to determine properties
of the integrable model, but it might be possible to study the
analytic structure of the free energy \(F_0\) to gain some
understanding of the Bethe roots \(\bm{x}\). This problem is currently
under further investigation, and we hope to report new results
elsewhere.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{KP tau function}
Lastly, one may also observe that if we fix one set of coordinates,
say \(\bm{y}\), then we can write
\begin{equation}
 \mathcal{I}(N,M|\bm{t}) 
 = \sum_{\lambda} s_\lambda(\bm{y})  s_\lambda(\bm{t}) 
 \equiv \sum_{\lambda} c_\lambda(\bm{y})  s_\lambda(\bm{t}) \; ,
\end{equation}
with coefficients \(c_\lambda(\bm{y}) = \det (h_{\lambda_i-i
  +j}(\bm{y}))\). But now, it is trivial to notice that these are
Plücker coordinates in the Jacobi-Trudi form, as seen
in~\cite{Miwa2000, Alexandrov:2012tr}.

Hence, the following expression
\begin{equation}
 \mathcal{I}(N,M|\bm{t}) = \sum_{\lambda \subseteq [N,M]} s_\lambda(\bm{y})  s_\lambda(\bm{t}) \; ,
\end{equation}
is also a KP tau function, a fact that we already know
from~\cite{Wheeler:2010vmq}, where the author proved this statement
using the free fermions formalism.

@@

\subsection{Correlation functions}
Bogoliubov has also shown in~\cite{Bogoliubov2005}, that the
correlation functions
\begin{subequations}
\begin{equation}
\begin{split}
\label{eq:correlation}
  A_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\})
  & = \bra{0} \prod_{j=1}^N \mathbb{C}(x_j)
  \prod_{k=1}^{N-1} \mathbb{B}(y_k) \phi_m^\dagger \ket{0}\\
  & =  \prod_{j=1}^N x_{j}^{M/2} \prod_{k=1}^{N-1} y_{j}^{M/2}
  \bra{0} \prod_{j=1}^N C(1/x_j) \prod_{k=1}^{N-1} B(y_k) \phi_m^\dagger \ket{0}
\end{split}
\end{equation}
can be written as
\begin{equation}
  A_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\}) = 
  \frac{(-1)^{N-1}}{y_N^{(N-1)/2}} \prod_{j=1}^N x_{j}^{M/2}
  \prod_{k=1}^{N-1} y_{j}^{M/2}
  \left( \prod_{t<N} \frac{y_N - y_t}{y_t} \right)
  \frac{\det Q}{\det H} \mathcal{I}(N,M|\bm{x}, \bm{y})
\end{equation}
\end{subequations}
where \(Q\) is an \(N\times N\) matrix with components 
\begin{equation}
 Q_{jN} = x_j^{(M + N - 1- 2m)/2} \quad  \textrm{and} \quad 
 Q_{jk} = H_{jk} \; , 
\end{equation}
and \(H_{jk} = H(x_j, y_k)\) are the components of the matrix \(H\)
in~(\ref{eq:h-matrix}). Since the components \(Q_{jN}\) are
independent of the coordinates \(y\), we cannot express the above
expression as a Toda hierarchy tau function.
